Next Trippy Structures\thanks{Supported by organization x.}

### 2019-07-11 paper.tex
Rev Daniil v3

\section{Related works}
\label{sec:rw}
\subsection{Trajectory indexing}
Literature on spatial trajectory indexing can be categorized by the nature of the trajectories: they can be either constrained to a network or in free space. While there are well-known queries for indexes that work on moving objects in free space \cite{DBLP:conf/vldb/PfoserJT00}, the network constrained trajectory indexes cover more diverse querying needs, as different networks involve different kinds of challenges (as in vehicular road network vs public transportation network), and also because the intended application may vary (analyzing trip-chaining patterns vs number of passengers within a time window). A comprehensive review on indexing methods can be found in \cite[Chapter 4]{DBLP:books/sp/PelekisT14}, to which we shall expand in the rest of this section to mention the most notable indexes and some new developments \ojo{Desde la referencia NO}.

\subsubsection{Free trajectory indexing}
Several adaptations of the {\em R-Tree} \cite{DBLP:conf/sigmod/Guttman84} are widely used for the indexing of moving objects. The most common approach is to integrate the temporal dimension in the R-Tree itself, as found in the {\em STR-Tree} and the {\em TB-Tree} from \cite{DBLP:conf/vldb/PfoserJT00}.
Another common approach is to complement the R-Tree with another similar structure for time representation, as has been done for the {\em MV3R-tree}~\cite{DBLP:conf/vldb/PapadiasT01},
where an Historical {\em R-Tree}~\cite{nascimento1998towards} to partition on the temporal dimension.

@@

Recently a framework based on Apache Spark called {\em UlTraMan}~ \cite{ding2018ultraman} was developed. It supports different kinds of partitioning schemes for large collections of trajectories to answer range, kNN or aggregation queries \ojo{Queries NO definidas aun}, allowing to divide the dataset to maximize the query time efficiency for a given query type. \ojo{Nieves: No claro} Although UlTraMan has been tested over network constrained trajectories, it does not appear to be exploiting network information in any way, hence its inclusion in this category.

@@

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels. \ojo{Nieves: no claro} Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of moving objects \ojo{NO} along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives, such as the {\em TMN-Tree}~\cite{chang2010tmn}, integrate {\em B$^+$-Trees}, which have proven to be more space and time efficient for indexing the temporal dimension. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

@@

Another relevant representation is described in \cite{DBLP:conf/gis/KroghPTT14}. Their proposed index, {\em NETTRA}, was designed to efficiently solve a specific kind of query called {\em Strict Path Queries (SPQ)} that ensure that selected trajectories not only touch with the searched path but also follow it from its beginning to its end. Built on a traditional RDBMS with $B^+$-Tree indexes. Another distinctive feature of NETTRA is its treatment of network constrained trajectories as textual information, which allows to apply string matching techniques such as fingerprinting functions to determine what trajectories have similar paths on their traversed edges. This close equivalence between trajectories and strings has been further exploited by Geodabs~\cite{chapuis2018geodabs}, where both the spatial distribution and sequence information are taken into account for finding trajectories by similarity with fingerprinting. These text-based approaches are sometimes tangled with works on the topic of semantic trajectories such as \cite{al2017semantictraj}.

@@

A recent compact data structure named {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. Due to space optimization, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in street networks, it cannot be applied to our problems in public transportation networks. \ojo{¿Por que?}

@@

\begin{itemize}
    \item $rank_1(B,i)$ returns the number of set bits in $B[1..i]$. Alternatively, $rank_0(B,i) = i - rank_1(B,i)$ and also $B[i] = rank_1(B,i-1) - rank_1(B,i)$. \ojo{¿Que aporta?}
    \item $select_1(B,i)$ returns the position in $1..n$ where the $i$th 1 occurs. Therefore, $rank_1(B,select_1(B,i)) = i$.
\end{itemize}

@@

\subsubsection{Compressed Suffix Array (CSA)}
\label{sec:csa}
One common structure to count or locate all the occurrences of a pattern $P[1,m]$ in a string $S[1,n]$ of symbols from an alphabet $\Sigma$ is the Suffix Array (SA) \cite{MM93}. The SA lexicographically sorts all the suffixes $\{S[i..n] : 1 \leq i \leq n\}$ and builds an array $A[1,n]$ with their positions in the \ojo{???}, holding that $S[A[i]..n] \prec S[A[i+1]..n]$. This allows to find out all the occurrences of a substring $P[1,m]$ in $O(m \log n)$ time with binary searches.

A simple enhancement to avoid storing the original string $S$ is to align $A$ with a bitvector $D[1,n]$ where $D[i]=1~\Leftrightarrow~S[A[i-1]] \prec S[A[i]]$. That is, $D$ marks with a 1 all the positions in $A$ where the starting symbol of that suffix $S[A[i]]$ has a larger lexicographical value than the one from the previous positions $S[A[i-1]]$. With $D$, keeping $S$ is not longer needed as $S[A[i]] = rank_1(D,i)$. \ojo{Nieves no entiende}
%, where $rank_1(D,i)$ returns the number of set bits in $D[1..i]$.

@@

\section{Our model for compact trip representation}
\label{sec:model}
In our previous structures introduced in \cite{brisaboa2018new}, we propose a model to represent the demand information for a public transportation network, as seen in the Figure~\ref{fig:er}. \ojo{Este parrafo regu}

@@

\begin{enumerate}[A)]
    \item Queries about the network load, asking for the gross number of users that boarded or alighted within a stop and a time/journey. Furthermore, it can be also interesting to obtain the average load of a bus or a train between any two stops from its line. Some of those queries are:
    \begin{itemize}
        \item \texttt{board\_X$_{LT}$} Number of users that stop X is boarded, optionally restricting to a line L and a time range T.
        \item \texttt{alight\_X$_{LT}$} Number of users that stop X is alighted, optionally restricting to a line L and a time range T.
        \item \texttt{board\_L$_T$}  Number of users that line L is boarded, optionally restricting to a time range T.
        \item \texttt{alight\_L$_T$} Number of users that line L is alighted, optionally restricting to a time range T.
        \item \texttt{board\_T} Number of users boarding within a time range T.
        \item \texttt{alight\_T} Number of users alighting within a time range T.
        \item \texttt{load\_X$_{LT}$} The average number of passengers traveling from the stop X to its next stop in the line L within the time range T. Can also be seen as the average load of the vehicle.
    \end{itemize}
    \item Queries about user trips patterns. With his kind of queries we can obtain the number of times a stop was used to switch lines or the number of trips that started on a stop with another specific stop as the final destination. In this work we consider the following queries of this kind:
    \begin{itemize}
        \item \texttt{start\_X$_{LT}$} User trips starting at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{end\_X$_{LT}$} User trips ending at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{switch\_X$_{L\textsubscript{1}L\textsubscript{2}T}$} Number of trips in which the stop X was used to switch lines, optionally restricting to a line L\textsubscript{2} and a time range T.
        \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} User trips that originate at stop X to end at stop Y, both being optionally restricted to a line and time range.
        \item \texttt{start\_L$_T$} User trips starting at any stop from the line L, optionally restricting to a time range T.
        \item \texttt{end\_L$_T$} User trips ending at any stop from the line L, optionally restricting to a time range T. \ojo{revisar}
        \item \texttt{start\_T} User trips starting within a time range T.
        \item \texttt{end\_T} User trips ending within a time range T.
    \end{itemize}
\end{enumerate}

\section{Proposed Structures}
\label{sec:ps}
To the best of our knowledge, there is no indexing structure that would allow to efficiently represent trajectories that could also support all the kinds of queries described in Section~\ref{sec:rq}. For this reason, we propose two data structures, \ojo{UNA SOLA SOLUCION} \acummname~(\acumm)~and~\ttctrname~(\ttctr). The former is targeted for queries of type A, solving most aggregation queries in constant time, while the later can be used for queries of type B. Finally, we introduce a more versatile alternative to \ttctr~that we call \ctrname~(\ctr).

\subsection{Common Data Structures}
\label{sec:cs}
\ojo{TRATAR COMO UNA SOLA SOLUCION} Both \acumm~and \ttctr~need a network representation to be operated, along with information about the transport vehicle journeys. Therefore, the following structures represent a network with stops $s_i \in S = \{1,...|S|\}$, lines $l_i \in L = \{1,...|L|\}$ and journeys $j_i \in J^l = \{1,...|J^l|\}$. It is important to state that journeys are \textbf{not} identified by $j_i$, as the same $j_i$ can belong to several $J^l$ from different lines, so we speak about journey \textbf{codes} (jcodes) instead of journey identifiers.

@@

\subsection{\acumm}
\label{sec:acumm}
As it has been stated previously in this paper, nowadays it could be relatively simple to gather massive information about user trips in every public transport; however, it is not necessary to store individually each trip. If there are two (or more) traveler sharing the same bus (train/tram/etc), they are also sharing the route, the stops (the time of each stop) and even the network. Therefore, a structure storing the aggregated data of the user trips would be a handy solution to solve (at least) queries of type A presented in section 3.

@@

A more complex \ojo{mas que que?} query that we can solve with \ctr~is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window (\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} in Section~\ref{sec:rq}). The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it works as the $range$ operation defined in Section~\ref{sec:wm}, also reporting the range in the last level of a WM corresponding to the entries of a single queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient).

@@

Of course, \ctr~can also be used to efficiently obtain other interesting information about trips, such as the top k most boarded stops, as shown in \cite{brisaboa2018compact} \ojo{NO}, with the additional possibility of differentiating stops that are only used to switch lines in \ctr. However, to the best of our knowledge, there is no known way of using this representation to obtain other kinds of information efficiently (e.g. the number of passengers in a journey between two stops).

\subsection{Analyzing \acumm, \ttctr~and \ctr}
\ojo{Mucho espacio}
\label{sec:comp}
To illustrate the application of each representation, as well as highlight the motivation for the development of \ctr, in this section we discuss in detail the worst case time complexities of each query described in Section~\ref{sec:rq}.

@@

\begin{threeparttable}
\centering
\caption{Worst case time complexities of the representations in Section~\ref{sec:ps}}
\label{tab:queries}
\begin{tabular}{|l|l|l|l|}
\hline
Query &  \acumm & \ttctr & \ctr\\
\hline
\texttt{board\_X$_{LT}$} & $O(l)$ & $O(l\times \log(\bar{|J^l|}))$\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{alight\_X$_{LT}$} & $O(l)$ & Hard\tnote{$\ddagger$$\diamondsuit$} & Hard\tnote{$\ddagger$$\diamondsuit$} \\
\texttt{board\_L$_T$} & O(1) & $O(|S|\times \log(\bar{|J^l|}))$\tnote{$\otimes$} \hspace{0.4cm} & $O(|S| \times \ctrq)$\tnote{$\otimes$} \\
\texttt{alight\_L$_T$} & O(1) & $O(|S|\times \log(\bar{|J^l|}))$\tnote{$\otimes$$\ddagger$} & $O(|S| \times \ctrq)$\tnote{$\otimes$$\ddagger$} \\
\texttt{board\_T} & $O(|L|)$ & Hard\tnote{$\diamondsuit$} & Hard\tnote{$\diamondsuit$} \\
\texttt{alight\_T} & $O(|L|)$ & Hard\tnote{$\ddagger$$\diamondsuit$} & Hard\tnote{$\ddagger$$\diamondsuit$} \\
\texttt{load\_X$_{LT}$} & $O(l)$ & Hard\tnote{$\ddagger$$\diamondsuit$} & Hard\tnote{$\ddagger$$\diamondsuit$} \\
\hline
\texttt{start\_X$_{LT}$} & - & \ttctrq\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{end\_X$_{LT}$} & - & $O(|S|\times \ttctrq)$\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{switch\_X$_{LT}$} & - & \ttctrq\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} & - & $O(|S|\times \ttctrq)$\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{start\_L$_T$} & - & $O(|S|\times \ttctrq)$\tnote{$\otimes$}~~~ & $O(\log(|L|\times\bar{|J^l|}))$ \\
\texttt{end\_L$_T$} & - & Hard\tnote{$\diamondsuit$} & $O(|S| \times \ctrq)$\tnote{$\otimes$} \\
\texttt{start\_T} & - & Hard\tnote{$\diamondsuit$} & $O(|L| \times \log(|L|\times\bar{|J^l|}))$~~ \\
\texttt{end\_T} & - & Hard\tnote{$\diamondsuit$} & Hard\tnote{$\diamondsuit$} \\
\hline
\end{tabular}

\ojo{Dar visto bueno}

@@

    \item \texttt{board\_L$_T$} Straightforward in \acumm, as it only needs to access one matrix. For \ttctr~we must filter through the \texttt{WM} for every stop from L. While in \ctr~the total number of occurrences of line L can be calculated in one $O(\log|L|)$ operation (with additional filtering through \FuncSty{WMJ} if needed), we need to subtract the occurrences of start\_L$_T$ and end\_L$_T$, the later having a greater complexity of $O(|S| \times \ctrq)$.
    \ojo{Consultas sin sentido}
    \item \texttt{alight\_L$_T$} Can be approximated both in \ttctr~and \ctr~with board\_L$_T$, given that every user that boards on a line is expected to alight from it within a reasonable time.
    \item \texttt{board\_T} and \texttt{alight\_T} Although \acumm, \ttctr~and \ctr~could solve it by applying board\_L$_T$ $|L|$ times, this would imply a prohibitive cost for both \ttctr~and \ctr, where extracting all the trips would be preferred in practice.

>>>>

\section{Related works}
\label{sec:rw}
\subsection{Trajectory indexing}
Literature on spatial trajectory indexing can be categorized by the nature of the trajectories: they can be either constrained to a network or in free space. While there are well-known queries for indexes that work on moving objects in free space \cite{DBLP:conf/vldb/PfoserJT00}, the network constrained trajectory indexes cover more diverse querying needs, as different networks involve different kinds of challenges (as in vehicular road network vs public transportation network), and also because the intended application may vary (analyzing trip-chaining patterns vs number of passengers within a time window). A comprehensive review on indexing methods can be found in \cite[Chapter 4]{DBLP:books/sp/PelekisT14}.
%, to which we shall expand in the rest of this section to mention the most notable indexes and some new developments \ojo{Desde la referencia NO}.

\subsubsection{Free trajectory indexing}
Several adaptations of the {\em R-Tree} \cite{DBLP:conf/sigmod/Guttman84} are widely used for the indexing of moving objects. The most common approach is to integrate the temporal dimension in the R-Tree itself, as found in the {\em STR-Tree} and the {\em TB-Tree} from \cite{DBLP:conf/vldb/PfoserJT00}.
Another common approach is to complement the R-Tree with another similar structure for time indexing, as has been done for the {\em MV3R-tree}~\cite{DBLP:conf/vldb/PapadiasT01},
where an Historical {\em R-Tree}~\cite{nascimento1998towards} to partition on the temporal dimension.

@@

Recently a framework based on Apache Spark called {\em UlTraMan}~ \cite{ding2018ultraman} was developed. It supports different kinds of partitioning schemes for large collections of trajectories to answer range, kNN or aggregation queries \ojo{Queries NO definidas aun}, allowing to dynamically change the partitioning of the dataset in order to maximize the query time efficiency for a given query type. Although UlTraMan has been tested over network constrained trajectories, it does not appear to be exploiting network information in any way, hence its inclusion in this category.

@@

\subsubsection{Network constrained trajectory indexing}
There are also numerous approaches that use R-Tree based indexes for trajectories that are constrained to an underlying network, aiming to decrease the redundancy in the representation by separating the representation into levels: the network elements (usually roads or road segments), and the trajectories over this network which, unlike the network elements, is expected to grow asymptotically over time. Examples of such indexes include the {\em FNR-Tree}~\cite{DBLP:conf/ssd/Frentzos03}, where the network elements are indexed with a 2D R-Tree and 1D R-Trees to index time intervals of the objects' movements along the edges of the network. Some of the limitations of the FNR-Tree are addressed in \cite{DBLP:journals/geoinformatica/AlmeidaG05}, where the authors propose the {\em MON-Tree}, where the moving objects are indexes in two dimensional R-Trees (the dimensions being edge position vs time). More recent alternatives, such as the {\em TMN-Tree}~\cite{chang2010tmn}, integrate {\em B$^+$-Trees}, which have proven to be more space and time efficient for indexing the temporal dimension. Alternatively, in \cite{rivera2018faster} a compact representation of time intervals is proposed using two bitvectors, that can be combined with these R-Tree based indexes to increase the efficiency for temporal filtering. Refer to \cite{john2017performance} for a quick comparison of these R-Tree based indexes.

@@

Another relevant representation is described in \cite{DBLP:conf/gis/KroghPTT14}. Their proposed index, {\em NETTRA}, was designed to efficiently solve a specific kind of query called {\em Strict Path Queries (SPQ)} that finds all the trajectories that follow the queried path from the beginning to its end. Built on a traditional RDBMS with $B^+$-Tree indexes. Another distinctive feature of NETTRA is its treatment of network constrained trajectories as textual information, which allows to apply string matching techniques such as fingerprinting functions to determine what trajectories have similar paths on their traversed edges. This close equivalence between trajectories and strings has been further exploited by Geodabs~\cite{chapuis2018geodabs}, where both the spatial distribution and sequence information are taken into account for finding trajectories by similarity with fingerprinting. These text-based approaches are sometimes tangled with works on the topic of semantic trajectories such as \cite{al2017semantictraj}.

@@

A recent compact data structure named {\em CiNCT} has been proposed in \cite{koide2018cinct}, where trajectories are encoded into a string, that is used to build an FM-index \cite{DBLP:conf/focs/FerraginaM00} with a Huffman-shaped Wavelet Tree \cite{ferragina2009compressed}. Due to space optimization, the string is constructed with relative movement labels instead of absolute edges, with an auxiliary structure that represents a network graph built from the input trajectories themselves. While this structure excels at pattern-matching and path extraction in street networks, it is not the most adequate for studying traveling patterns over public transportation networks that are based on routes, stops or similar abstractions.

@@

\begin{itemize}
    \item $rank_1(B,i)$ returns the number of set bits in $B[1..i]$. Alternatively, $rank_0(B,i) = i - rank_1(B,i)$ and also $B[i] = rank_1(B,i-1) - rank_1(B,i)$.
    \item $select_1(B,i)$ returns the position in $1..n$ where the $i$th 1 occurs. Therefore, $rank_1(B,select_1(B,i)) = i$.
\end{itemize}

@@

\subsubsection{Compressed Suffix Array (CSA)}
\label{sec:csa}
One common structure to count or locate all the occurrences of a pattern $P[1,m]$ in a string $S[1,n]$ of symbols from an alphabet $\Sigma$ is the Suffix Array (SA) \cite{MM93}. The SA is an array $A[1,n]$ with the positions $i$ of all suffixes $\{S[i..n] : 1 \leq i \leq n\}$, lexicographically sorted holding that $S[A[i]..n] \prec S[A[i+1]..n]$. This allows to find out all the occurrences of a substring $P[1,m]$ in $O(m \log n)$ time with binary searches.

A simple enhancement to avoid storing the original string $S$ is to align $A$ with a bitvector $D[1,n]$ where $D[i]=1~\Leftrightarrow~S[A[i-1]] \prec S[A[i]]$. That is, $D$ marks with a 1 all the positions in $A$ where the starting symbol of that suffix $S[A[i]]$ has a larger lexicographical value than the one from the previous positions $S[A[i-1]]$. With $D$, keeping $S$ is not longer needed as $S[A[i]] = rank_1(D,i)$.
%, where $rank_1(D,i)$ returns the number of set bits in $D[1..i]$.

@@

\section{Our model for compact trip representation}
\label{sec:model}
In our previous representation introduced in \cite{brisaboa2018new}, we propose a model to represent the demand information for a public transportation network, as seen in the Figure~\ref{fig:er}.

@@

\begin{enumerate}[A)]
    \item Queries about the network load, asking for the gross number of users that boarded or alighted within a stop and a time/journey. Furthermore, it can be also interesting to obtain the average load of a bus or a train between any two stops from its line. Some of those queries are:
    \begin{itemize}
        \item \texttt{board\_X$_{LT}$} Number of users that stop X is boarded, optionally restricting to a line L and a time range T.
        \item \texttt{alight\_X$_{LT}$} Number of users that stop X is alighted, optionally restricting to a line L and a time range T.
        \item \texttt{use\_L$_T$}  Number of users (boardings) for the line L, optionally restricting to a time range T.
        \item \texttt{board\_T} Number of users boarding within a time range T.
        \item \texttt{alight\_T} Number of users alighting within a time range T.
        \item \texttt{load\_X$_{LT}$} The average number of passengers traveling from the stop X to its next stop in the line L within the time range T. Can also be seen as the average load of the vehicle.
    \end{itemize}
    \item Queries about user trips patterns. With his kind of queries we can obtain the number of times a stop was used to switch lines or the number of trips that started on a stop with another specific stop as the final destination. In this work we consider the following queries of this kind:
    \begin{itemize}
        \item \texttt{start\_X$_{LT}$} User trips starting at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{end\_X$_{LT}$} User trips ending at a stop X, optionally restricting to a line L and a time range T.
        \item \texttt{switch\_X$_{L\textsubscript{1}L\textsubscript{2}T}$} Number of trips in which the stop X was used to switch lines, optionally restricting to a line L\textsubscript{2} and a time range T.
        \item \texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} User trips that originate at stop X to end at stop Y, both being optionally restricted to a line and time range.
        \item \texttt{start\_L$_T$} User trips starting at any stop from the line L, optionally restricting to a time range T.
        \item \texttt{end\_L$_T$} User trips ending at any stop from the line L, optionally restricting to a time range T. \ojo{revisar}
        \item \texttt{start\_T} User trips starting within a time range T.
        \item \texttt{end\_T} User trips ending within a time range T.
    \end{itemize}
\end{enumerate}

\section{Proposed Structures}
\label{sec:ps}
To the best of our knowledge, there is no indexing structure that would allow to efficiently represent trajectories that could also support all the kinds of queries described in Section~\ref{sec:rq}. For this reason, we propose a solution that relies on two data structures, \acummname~(\acumm)~and~\ttctrname~(\ttctr). The former is targeted for queries of type A, solving most aggregation queries in constant time, while the later can be used for queries of type B. Finally, we introduce a more versatile alternative to \ttctr~that we call \ctrname~(\ctr).

\subsection{Common Data Structures}
\label{sec:cs}
Our developed solution needs a network representation to be operated, along with information about the transport vehicle journeys. Therefore, the following structures represent a network with stops $s_i \in S = \{1,...|S|\}$, lines $l_i \in L = \{1,...|L|\}$ and journeys $j_i \in J^l = \{1,...|J^l|\}$. It is important to state that journeys are \textbf{not} identified by $j_i$, as the same $j_i$ can belong to several $J^l$ from different lines, so we speak about journey \textbf{codes} (jcodes) instead of journey identifiers.

@@

\subsection{\acumm}
\label{sec:acumm}
As it has been stated previously in this paper, nowadays it could be relatively simple to gather massive information about user trips in every public transport; however, it is not necessary to store individually each trip. If there are two (or more) travelers sharing the same bus (train/tram/etc), they are also sharing the route, the stops (the time of each stop) and even the network. Therefore, a structure storing the aggregated data of the user trips would be a handy solution to solve (at least) queries of type A presented in section 3.

@@

An example of a complex query that we can solve with \ctr~is ``number of trips that started from a stop \DataSty{s$_a$} and ended at a stop \DataSty{s$_z$}'', which can be further restricted to specific starting and ending lines and a time window (\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} in Section~\ref{sec:rq}). The pseudocode for the full version of such query can be found at the Algorithm~\ref{alg:xy}. This algorithm relies heavily on the abstract function \FuncSty{GetRange}, which when applied to a CSA, delimits the range of entries of $\Psi$ that match a queried string pattern, while on a WM it works as the $range$ operation defined in Section~\ref{sec:wm}, also reporting the range in the last level of a WM corresponding to the entries of a single queried symbol within a range in the encoded sequence (equivalent to two \FuncSty{TrackDown} operations on the first and last occurrence of the queried symbol, but more time efficient).

@@

Of course, \ctr~can also be used to efficiently obtain other interesting information about trips, such as the top k most boarded stops, 
%as shown in \cite{brisaboa2018compact} \ojo{NO}, 
with the possibility of differentiating stops that are only used to switch lines in \ctr. However, to the best of our knowledge, there is no known way of using this representation to obtain other kinds of information efficiently (e.g. the number of passengers in a journey between two stops).

\subsection{Analyzing \acumm, \ttctr~and \ctr}
\label{sec:comp}
To illustrate the application of each representation, as well as highlight the motivation for the development of \ctr, in this section we discuss in detail the worst case time complexities of each query described in Section~\ref{sec:rq}.

@@

\begin{threeparttable}
\centering
\caption{Worst case time complexities of the representations in Section~\ref{sec:ps}}
\label{tab:queries}
\begin{tabular}{|l|l|l|l|}
\hline
Query &  \acumm & \ttctr & \ctr\\
\hline
\texttt{board\_X$_{LT}$} & $O(l)$ & $O(l\times \log(\bar{|J^l|}))$\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{alight\_X$_{LT}$} & $O(l)$ & Hard\tnote{$\ddagger$$\diamondsuit$} & Hard\tnote{$\ddagger$$\diamondsuit$} \\
\texttt{use\_L$_T$} & O(1) & $O(|S|\times \log(\bar{|J^l|}))$\tnote{$\otimes$} \hspace{0.4cm} & $O(|S| \times \ctrq)$\tnote{$\otimes$} \\
%\texttt{alight\_L$_T$} & O(1) & $O(|S|\times \log(\bar{|J^l|}))$\tnote{$\otimes$$\ddagger$} & $O(|S| \times \ctrq)$\tnote{$\otimes$$\ddagger$} \\
\texttt{board\_T} & $O(|L|)$ & Hard\tnote{$\diamondsuit$} & Hard\tnote{$\diamondsuit$} \\
\texttt{alight\_T} & $O(|L|)$ & Hard\tnote{$\ddagger$$\diamondsuit$} & Hard\tnote{$\ddagger$$\diamondsuit$} \\
\texttt{load\_X$_{LT}$} & $O(l)$ & Hard\tnote{$\ddagger$$\diamondsuit$} & Hard\tnote{$\ddagger$$\diamondsuit$} \\
\hline
\texttt{start\_X$_{LT}$} & - & \ttctrq\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{end\_X$_{LT}$} & - & $O(|S|\times \ttctrq)$\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{switch\_X$_{LT}$} & - & \ttctrq\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{from\_X$_{LT}$\_to\_Y$_{LT}$} & - & $O(|S|\times \ttctrq)$\tnote{$\otimes$} & \ctrq\tnote{$\otimes$} \\
\texttt{start\_L$_T$} & - & $O(|S|\times \ttctrq)$\tnote{$\otimes$}~~~ & $O(\log(|L|\times\bar{|J^l|}))$ \\
\texttt{end\_L$_T$} & - & Hard\tnote{$\diamondsuit$} & $O(|S| \times \ctrq)$\tnote{$\otimes$} \\
\texttt{start\_T} & - & Hard\tnote{$\diamondsuit$} & $O(|L| \times \log(|L|\times\bar{|J^l|}))$~~ \\
\texttt{end\_T} & - & Hard\tnote{$\diamondsuit$} & Hard\tnote{$\diamondsuit$} \\
\hline
\end{tabular}

@@

    \item \texttt{use\_L$_T$} Straightforward in \acumm, as it only needs to access one matrix. For \ttctr~we must filter through the \texttt{WM} for every stop from L. While in \ctr~the total number of occurrences of line L can be calculated in one $O(\log|L|)$ operation (with additional filtering through \FuncSty{WMJ} if needed), we need to subtract the occurrences of start\_L$_T$ and end\_L$_T$, the later having a greater complexity of $O(|S| \times \ctrq)$.
    %\item \texttt{alight\_L$_T$} Can be approximated both in \ttctr~and \ctr~with board\_L$_T$, given that every user that boards on a line is expected to alight from it within a reasonable time.
    \item \texttt{board\_T} and \texttt{alight\_T} Although \acumm, \ttctr~and \ctr~could solve it by applying use\_L$_T$ $|L|$ times, this would imply a prohibitive cost for both \ttctr~and \ctr, where extracting all the trips would be preferred in practice.
